Finite-state morphological transducers for three Kypchak languages

### 2014-03-14 kipchak-lrec-2013/kipchak-paper.tex
irregular harmony
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@50751 72bbbca6-d526-0410-a7d9-f06f51895060

\defaultfontfeatures{Mapping=tex-text}
\setromanfont{Times New Roman}
\newfontfamily\symbl[Scale=MatchLowercase]{FreeSerif}

@@

%FIXME: description of each one, with examples
\begin{itemize}
 \item \textbf{Ambiguous and `irregular' harmony}: In Tatar, certain characters in the alphabet which represent vowels can be 
   ambiguous between a vowel and a `yoticised' vowel, such as е, which is front or back. For 
   example: дәрес `lesson', дәресләр `lessons'; ел `year', еллар `years'. 
 \item \textbf{Loan words}: In Kumyk, the character ё in Russian words is pronounced as in Russian where in native words 
   it represents the sound /ø/. These two pronunciations will select for different suffixes, the Russian pronunciation taking
   back vowel suffixes самолётлар `aeroplanes', and /ø/ taking front vowel suffixes сёзлер `words'.
 \item \textbf{Acronyms and numerals}: Acronyms and numerals are challenging as they will often be pronounced out loud
   in their non-abbreviated forms, for example in Kazakh, 30-дан `from thirty', 5-тен `from 5'. The ablative
   suffix -DAn alters for the phonology, but in the numeral string there is no indication of how they should alter.
\end{itemize}

@@

% FIXME: cyrillic in \texttt
Our solution is to add phonological information at the end of morphemes that need it in the form of special ``abstract letters'' that trigger phonological processes at the morphophonological stage and are deleted. For example, the string \texttt{5<num><subst><abl>}\footnote{The resulting forms, 5-тен\super{\texttt{kaz}}, 5-тән\super{\texttt{tat}}, and 5-ден\super{\texttt{kum}} would be pronounced бестен\super{\texttt{kaz}}, биштән\super{\texttt{tat}}, and бешден\super{\texttt{kum}}.} would have
the morphotactic representation 5-\{э\}\{с\}>\{D\}\{A\}н, where \{э\} and \{с\} stand for phonological triggers, and \{D\}\{A\}н is the representation of the ablative morpheme at the morphotactic level. The symbol \{э\} signals that the following vowel needs to harmonise to a front unrounded vowel, and \{с\} signals that there is a final voiceless consonant.  So, the rules that operate on \{D\} and \{A\} will be able to do the forward voicing assimilation \{D\} surfacing as т /t/ and the front harmony \{A\} surfacing as е /e/.

>>>>

\defaultfontfeatures{Mapping=tex-text}
\setromanfont{Times New Roman}
\newfontfamily\symbl[Scale=MatchLowercase]{FreeSerif}
\newfontfeature{IPA}{+mgrk}
\newfontfamily\qipa[IPA,Scale=MatchLowercase]{FreeSerif}

@@

%FIXME: description of each one, with examples
\begin{description}
 \item{\textbf{Ambiguous and `irregular' harmony}:} In Tatar, the `yoticised' vowel letters ‹е›, ‹я›, and ‹ю› each ambiguously represent a set of sounds.  For example, ‹е› can represent the sounds of /j/ followed the sound of ‹э› (as in ‹егетләр› `boys') or ‹ы› (as in ‹еллар› `years'), and is also the non-initial orthographic variant of ‹э› (as in ‹дәресләр› `lessons').  In Kumyk, ‹ё› and ‹ю› are used intervocalically to represent front rounded vowels (as in ‹гёзлер› `eyes' and ‹гюнлер› `days'), but word-initially can represent ‹й› followed by either a front (as in ‹юреклер› `hearts', ‹ёнкюлер› `darlings') or a back (as in ‹юлдузлар› `stars', ёллар `roads') vowel.  In Kazakh, the vowels /{\qipa ɘ}/ ‹і› and /ə/ ‹ы› followed by a glide /w/ ‹у› or /j/ ‹й› are written ‹у› and ‹и›, depending on the glide; for example.  The character ‹ю› is used in turn to represent ‹й› followed by either multi-phoneme value of ‹у›.  %FIXME: good Kazakh examples 
 \item \textbf{Loan words}: In Kumyk, the character ё in Russian words is pronounced as in Russian where in native words 
   it represents the sound /ø/. These two pronunciations will select for different suffixes, the Russian pronunciation taking
   back vowel suffixes самолётлар `aeroplanes', and /ø/ taking front vowel suffixes сёзлер `words'.  %FIXME: -ия in Tatar
 \item \textbf{Acronyms and numerals}: Acronyms and numerals are challenging as they will often be pronounced out loud
   in their non-abbreviated forms, for example in Kazakh, 30-дан `from thirty', 5-тен `from 5'. The ablative
   suffix -DAn alters for the phonology, but in the numeral string there is no indication of how they should alter.
\end{description}

@@

% FIXME: cyrillic in \texttt
Our solution is to add phonological information at the end of morphemes that need it in the form of special ``abstract letters'' that trigger phonological processes at the morphophonological stage and are deleted. For example, the string \texttt{5<num><subst><abl>}\footnote{The resulting forms, 5-тен\super{\texttt{kaz}}, 5-тән\super{\texttt{tat}}, and 5-ден\super{\texttt{kum}} would be pronounced бестен\super{\texttt{kaz}}, биштән\super{\texttt{tat}}, and бешден\super{\texttt{kum}}.} has
the morphotactic representation 5\{э\}\{с\}>-\{D\}\{A\}н, where \{э\} and \{с\} stand for phonological triggers, > represents a morpheme boundary, and \{D\}\{A\}н is the representation of the ablative morpheme at the morphotactic level.  The symbol \{э\} signals that the following vowel needs to harmonise to a front unrounded vowel, and \{с\} signals that there is a final voiceless consonant.  So, with 5\{э\}\{с\} in the lexicon, and a consistent continuation lexicon specifying the underlying form of case affixes, the rules that operate on \{D\} and \{A\} are able to do produce the correct output form in each language, avoiding the incorrect default *5-дан.
